---
title: "BELIEVE Phenotype Preparation"
author: "Joud Touqan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(jsonlite)
library(corrplot)
output_dir <- "/home/jt962/rds/hpc-work/believe_adiposity/01_phenotype_prep"
```

---

## 1. Read Data

```{r read-data}
phenotype_data_raw <- read.csv(
  "/home/jt962/rds/rds-post_qc_data-pNR2rM6BWWA/believe/phenotype/data_freeze_3/analysis_regeneron.csv",
  stringsAsFactors = FALSE)

cat("Participants:", nrow(phenotype_data_raw), "\n")
cat("Variables:",    ncol(phenotype_data_raw), "\n")
```

---

## 2. Select Columns

```{r select-columns}
phenotype_data <- phenotype_data_raw %>%
  select(genid, idno, study, ages, sex, ht, wt, bmi, waist, hip, whr, fatmass, fatperc)

cat("Selected:", nrow(phenotype_data), "rows x", ncol(phenotype_data), "columns\n")
```

---

## 3. Cleaning

```{r cleaning}
# Unique identifier
phenotype_data <- phenotype_data %>%
  mutate(unique_id = paste(study, idno, sep = "_"))

# Remove missing genid
n_no_genid <- sum(is.na(phenotype_data$genid))
phenotype_data <- phenotype_data %>% filter(!is.na(genid))
cat("Removed (no genid):", n_no_genid, "\n")

# Age filter
n_before_age <- nrow(phenotype_data)
phenotype_data <- phenotype_data %>% filter(ages >= 18)
cat("Removed (age < 18):", n_before_age - nrow(phenotype_data), "\n")
cat("Final analysis cohort:", nrow(phenotype_data), "\n")

# Sex coding — SAIGE: 0 = Male, 1 = Female
phenotype_data <- phenotype_data %>%
  mutate(sex_numeric = case_when(
    sex == "Male"   ~ 0,
    sex == "Female" ~ 1,
    TRUE            ~ NA_real_
  ))

cat("\nSex coding check:\n")
print(table(phenotype_data$sex, phenotype_data$sex_numeric, useNA = "always"))

# Age squared for covariate adjustment
phenotype_data <- phenotype_data %>% mutate(age_sq = ages^2)
```

---

## 4. Preliminary Exploration

### 4.1 Missing Data

```{r missing-data}
traits_check <- c("ht", "wt", "bmi", "waist", "hip", "whr", "fatmass", "fatperc")

missing_wide <- rbind(
  "N with data" = sapply(traits_check, function(v) sum(!is.na(phenotype_data[[v]]))),
  "N missing"   = sapply(traits_check, function(v) sum(is.na(phenotype_data[[v]])))
)
print(missing_wide)
```

### 4.2 Impossible Values

```{r impossible-values}
phenotype_data %>%
  summarise(across(c(bmi, waist, hip, whr, fatmass, fatperc),
                   ~ sum(. < 0, na.rm = TRUE),
                   .names = "{.col}_neg")) %>%
  print()
```

---

## 5. Outlier Removal

Values > 5 SDs from the mean are set to NA. Participant is retained.

```{r outliers}
flag_outliers <- function(x) {
  outlier <- abs(x - mean(x, na.rm = TRUE)) > 5 * sd(x, na.rm = TRUE)
  outlier[is.na(x)] <- FALSE
  return(outlier)
}

traits <- c("bmi", "waist", "hip", "whr", "fatmass", "fatperc", "ht", "wt")

outlier_summary <- data.frame(trait = traits, n_before = NA_integer_,
                               n_outliers = NA_integer_, n_after = NA_integer_)

for (i in seq_along(traits)) {
  trait   <- traits[i]
  n_before <- sum(!is.na(phenotype_data[[trait]]))
  outliers  <- flag_outliers(phenotype_data[[trait]])
  phenotype_data[[trait]][outliers] <- NA
  n_after  <- sum(!is.na(phenotype_data[[trait]]))
  outlier_summary[i, 2:4] <- c(n_before, n_before - n_after, n_after)
}

outlier_wide <- as.data.frame(t(outlier_summary[, -1]))
colnames(outlier_wide) <- outlier_summary$trait
rownames(outlier_wide) <- c("N before", "Outliers removed", "N after")
print(outlier_wide)
```

---

## 6. Pre-Adjustment Checks

### 6.1 Trait-to-Trait Correlation Matrix

```{r trait-correlation, fig.width=8, fig.height=7}
col_palette <- colorRampPalette(c("steelblue", "white", "firebrick"))(200)

cor_traits <- phenotype_data %>%
  select(bmi, waist, hip, whr, fatmass, fatperc) %>%
  cor(use = "pairwise.complete.obs")

cat("Trait correlation matrix (after cleaning):\n")
print(round(cor_traits, 3))

corrplot(cor_traits,
         method = "color", col = col_palette, type = "upper",
         addCoef.col = "black", number.cex = 1.1,
         tl.cex = 1.2, tl.col = "black",
         title = "Adiposity trait correlations (after cleaning)",
         mar = c(0, 0, 2, 0))

# Save PDF
pdf(file.path(output_dir, "corr_trait_to_trait.pdf"), width = 8, height = 7)
corrplot(cor_traits,
         method = "color", col = col_palette, type = "upper",
         addCoef.col = "black", number.cex = 1.1,
         tl.cex = 1.2, tl.col = "black",
         title = "Adiposity trait correlations (after cleaning)",
         mar = c(0, 0, 2, 0))
dev.off()
cat("Saved: corr_trait_to_trait.pdf\n")
```

### 6.2 Covariate vs Raw Trait Correlations (continuous covariates)

```{r cov-raw-corr, fig.width=9, fig.height=5}
cov_raw <- phenotype_data %>%
  select(ages, age_sq, sex_numeric, bmi, waist, hip, whr, fatmass, fatperc) %>%
  cor(use = "pairwise.complete.obs")

cov_vs_raw <- cov_raw[c("ages", "age_sq", "sex_numeric"),
                      c("bmi", "waist", "hip", "whr", "fatmass", "fatperc")]

cat("Covariate correlations with raw traits:\n")
print(round(cov_vs_raw, 3))

corrplot(cov_vs_raw,
         method = "color", col = col_palette,
         addCoef.col = "black", number.cex = 1.3,
         tl.cex = 1.2, tl.col = "black", is.corr = FALSE,
         title = "Covariate vs raw trait correlations (before adjustment)",
         mar = c(0, 0, 2, 0))
```

### 6.3 Study Site vs Raw Traits 

```{r site-raw-anova, fig.width=14, fig.height=10}
raw_traits <- c("bmi", "waist", "hip", "whr", "fatmass", "fatperc")

anova_raw <- data.frame(trait = raw_traits, F_stat = NA_real_,
                         p_anova = NA_real_, p_kruskal = NA_real_)

par(mfrow = c(2, 3), mar = c(6, 5, 4, 2), oma = c(0, 0, 2, 0))

for (i in seq_along(raw_traits)) {
  trait <- raw_traits[i]
  vals  <- phenotype_data[[trait]]
  grp   <- phenotype_data$study

  boxplot(vals ~ grp, col = c("steelblue","coral","lightgreen","mediumpurple"),
          main = trait, xlab = "", ylab = trait,
          las = 2, cex.main = 1.4, cex.lab = 1.2, cex.axis = 1.0)

  aov_res <- aov(vals ~ grp)
  f  <- summary(aov_res)[[1]][["F value"]][1]
  p  <- summary(aov_res)[[1]][["Pr(>F)"]][1]
  kp <- kruskal.test(vals ~ grp)$p.value

  mtext(paste0("F = ", round(f,1), "  p = ", signif(p,2)),
        side = 3, line = 0, cex = 0.9, col = "darkred")

  anova_raw[i, 2:4] <- c(round(f,2), signif(p,3), signif(kp,3))
}
mtext("Raw traits by study site (before adjustment)", outer = TRUE, cex = 1.3, font = 2)

cat("ANOVA and Kruskal-Wallis — raw traits by site:\n")
print(anova_raw)
```

### 6.4 Fat % by Sex (bimodal check)

```{r fatperc-sex, fig.width=10, fig.height=5}
par(mfrow = c(1, 2), mar = c(5, 5, 4, 2))

hist(phenotype_data$fatperc[phenotype_data$sex == "Male"],
     breaks = 40, col = "steelblue", border = "white",
     main = "Fat % - Male", xlab = "Fat %",
     xlim = range(phenotype_data$fatperc, na.rm = TRUE),
     cex.main = 1.4, cex.lab = 1.2, cex.axis = 1.1)

hist(phenotype_data$fatperc[phenotype_data$sex == "Female"],
     breaks = 40, col = "coral", border = "white",
     main = "Fat % - Female", xlab = "Fat %",
     xlim = range(phenotype_data$fatperc, na.rm = TRUE),
     cex.main = 1.4, cex.lab = 1.2, cex.axis = 1.1)
```


```{r save-preadj-pdfs}
# Covariate vs raw heatmap
pdf(file.path(output_dir, "corr_cov_vs_raw.pdf"), width = 9, height = 5)
corrplot(cov_vs_raw, method = "color", col = col_palette,
         addCoef.col = "black", number.cex = 1.3,
         tl.cex = 1.2, tl.col = "black", is.corr = FALSE,
         title = "Covariate vs raw trait correlations (before adjustment)",
         mar = c(0, 0, 2, 0))
dev.off()

# Boxplots — raw traits by site
pdf(file.path(output_dir, "boxplots_raw_by_site.pdf"), width = 14, height = 10)
par(mfrow = c(2,3), mar = c(6,5,4,2), oma = c(0,0,2,0))
for (trait in raw_traits) {
  boxplot(phenotype_data[[trait]] ~ phenotype_data$study,
          col = c("steelblue","coral","lightgreen","mediumpurple"),
          main = trait, xlab = "", ylab = trait,
          las = 2, cex.main = 1.4, cex.lab = 1.2, cex.axis = 1.0)
}
mtext("Raw traits by study site", outer = TRUE, cex = 1.3, font = 2)
dev.off()

# Fat % by sex
pdf(file.path(output_dir, "fatperc_by_sex.pdf"), width = 10, height = 5)
par(mfrow = c(1,2), mar = c(5,5,4,2))
hist(phenotype_data$fatperc[phenotype_data$sex == "Male"],
     breaks = 40, col = "steelblue", border = "white",
     main = "Fat % - Male", xlab = "Fat %",
     xlim = range(phenotype_data$fatperc, na.rm = TRUE),
     cex.main = 1.4, cex.lab = 1.2, cex.axis = 1.1)
hist(phenotype_data$fatperc[phenotype_data$sex == "Female"],
     breaks = 40, col = "coral", border = "white",
     main = "Fat % - Female", xlab = "Fat %",
     xlim = range(phenotype_data$fatperc, na.rm = TRUE),
     cex.main = 1.4, cex.lab = 1.2, cex.axis = 1.1)
dev.off()

cat("Pre-adjustment PDFs saved\n")
```

---

## 7. WHR-BMI Adjustment

```{r whr-adjustment}
whr_model <- lm(whr ~ bmi, data = phenotype_data, na.action = na.exclude)
phenotype_data$whr_bmi_adj <- residuals(whr_model)
cat("WHR adjusted for BMI. Non-missing:", sum(!is.na(phenotype_data$whr_bmi_adj)), "\n")
```

---

## 8. Covariate Adjustment

Age, age², sex, and study site regressed out of each trait.

```{r covariate-adjustment}
adjust_trait <- function(trait_col, data) {
  formula <- as.formula(paste(trait_col, "~ ages + age_sq + sex_numeric + study"))
  resids  <- residuals(lm(formula, data = data, na.action = na.exclude))
  return(resids)
}

traits_to_adjust <- c("bmi", "waist", "hip", "whr_bmi_adj", "fatmass", "fatperc")

adj_summary <- data.frame(trait = traits_to_adjust,
                           n_before = NA_integer_, n_after = NA_integer_)

for (i in seq_along(traits_to_adjust)) {
  trait   <- traits_to_adjust[i]
  adj_col <- paste0(trait, "_adj")
  adj_summary$n_before[i] <- sum(!is.na(phenotype_data[[trait]]))
  phenotype_data[[adj_col]] <- adjust_trait(trait, phenotype_data)
  adj_summary$n_after[i]  <- sum(!is.na(phenotype_data[[adj_col]]))
}

adj_wide <- as.data.frame(t(adj_summary[,-1]))
colnames(adj_wide) <- adj_summary$trait
rownames(adj_wide) <- c("N before", "N after")
print(adj_wide)
```

---

## 9. Inverse Normal Transformation (INT)

```{r int-transformation}
int_transform <- function(x) {
  ranks <- rank(x, na.last = "keep", ties.method = "average")
  qnorm((ranks - 0.5) / sum(!is.na(x)))
}

traits_adj <- paste0(traits_to_adjust, "_adj")

for (trait_adj in traits_adj) {
  int_col <- sub("_adj$", "_INT", trait_adj)
  phenotype_data[[int_col]] <- int_transform(phenotype_data[[trait_adj]])
}

cat("INT applied to all traits.\n")
```

---

## 10. Post-INT Checks

### 10.1 Covariate vs INT Trait Correlations

```{r cov-int-corr, fig.width=9, fig.height=5}
int_traits <- c("bmi_INT", "waist_INT", "hip_INT",
                "whr_bmi_adj_INT", "fatmass_INT", "fatperc_INT")

cov_int <- phenotype_data %>%
  select(ages, age_sq, sex_numeric, all_of(int_traits)) %>%
  cor(use = "pairwise.complete.obs")

cov_vs_int <- cov_int[c("ages", "age_sq", "sex_numeric"), int_traits]

cat("Covariate correlations with INT traits :\n")
print(round(cov_vs_int, 3))

corrplot(cov_vs_int,
         method = "color", col = col_palette,
         addCoef.col = "black", number.cex = 1.3,
         tl.cex = 1.2, tl.col = "black", is.corr = FALSE,
         title = "Covariate vs INT trait correlations (after adjustment and INT)",
         mar = c(0, 0, 2, 0))
```

### 10.2 Study Site vs INT Traits 

```{r site-int-anova, fig.width=14, fig.height=10}
anova_int <- data.frame(trait = int_traits, F_stat = NA_real_,
                         p_anova = NA_real_, p_kruskal = NA_real_)

par(mfrow = c(2,3), mar = c(6,5,4,2), oma = c(0,0,2,0))

for (i in seq_along(int_traits)) {
  trait <- int_traits[i]
  vals  <- phenotype_data[[trait]]
  grp   <- phenotype_data$study

  boxplot(vals ~ grp, col = c("steelblue","coral","lightgreen","mediumpurple"),
          main = trait, xlab = "", ylab = "INT value",
          las = 2, cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.0)

  aov_res <- aov(vals ~ grp)
  f  <- summary(aov_res)[[1]][["F value"]][1]
  p  <- summary(aov_res)[[1]][["Pr(>F)"]][1]
  kp <- kruskal.test(vals ~ grp)$p.value

  mtext(paste0("F = ", round(f,1), "  p = ", signif(p,2)),
        side = 3, line = 0, cex = 0.9, col = "darkred")

  anova_int[i, 2:4] <- c(round(f,2), signif(p,3), signif(kp,3))
}
mtext("INT traits by study site (after adjustment)", outer = TRUE, cex = 1.3, font = 2)

cat("ANOVA and Kruskal-Wallis — INT traits by site:\n")
print(anova_int)
```

### Save Post-INT PDFs

```{r save-postint-pdfs}
# Covariate vs INT heatmap
pdf(file.path(output_dir, "corr_cov_vs_int.pdf"), width = 9, height = 5)
corrplot(cov_vs_int, method = "color", col = col_palette,
         addCoef.col = "black", number.cex = 1.3,
         tl.cex = 1.2, tl.col = "black", is.corr = FALSE,
         title = "Covariate vs INT trait correlations (after adjustment)",
         mar = c(0, 0, 2, 0))
dev.off()

# Boxplots — INT traits by site
pdf(file.path(output_dir, "boxplots_int_by_site.pdf"), width = 14, height = 10)
par(mfrow = c(2,3), mar = c(6,5,4,2), oma = c(0,0,2,0))
for (trait in int_traits) {
  boxplot(phenotype_data[[trait]] ~ phenotype_data$study,
          col = c("steelblue","coral","lightgreen","mediumpurple"),
          main = trait, xlab = "", ylab = "INT value",
          las = 2, cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.0)
}
mtext("INT traits by study site (after adjustment)", outer = TRUE, cex = 1.3, font = 2)
dev.off()

cat("Post-INT PDFs saved\n")
```

---

## 11. Distribution Plots

```{r distribution-plots, fig.width=18, fig.height=12}
raw_18plus <- phenotype_data_raw %>%
  filter(ages >= 18, !is.na(genid))

trait_info <- list(
  list(raw=raw_18plus$bmi,     clean=phenotype_data$bmi,
       int=phenotype_data$bmi_INT,         name="BMI",      unit="kg/m2"),
  list(raw=raw_18plus$whr,     clean=phenotype_data$whr,
       int=phenotype_data$whr_bmi_adj_INT, name="WHR",      unit=""),
  list(raw=raw_18plus$waist,   clean=phenotype_data$waist,
       int=phenotype_data$waist_INT,       name="Waist",    unit="cm"),
  list(raw=raw_18plus$hip,     clean=phenotype_data$hip,
       int=phenotype_data$hip_INT,         name="Hip",      unit="cm"),
  list(raw=raw_18plus$fatmass, clean=phenotype_data$fatmass,
       int=phenotype_data$fatmass_INT,     name="Fat Mass", unit="kg"),
  list(raw=raw_18plus$fatperc, clean=phenotype_data$fatperc,
       int=phenotype_data$fatperc_INT,     name="Fat %",    unit="%")
)

#  PLOT FUNCTIONS 

plot_trait <- function(raw_vals, clean_vals, trait_name, unit) {
  outlier_mask <- flag_outliers(raw_vals)
  n_outliers   <- sum(outlier_mask, na.rm = TRUE)
  n_before     <- sum(!is.na(raw_vals))
  n_after      <- sum(!is.na(clean_vals))
  trait_mean   <- mean(clean_vals, na.rm = TRUE)

  hist(clean_vals,
       breaks   = 40,
       col      = "steelblue",
       border   = "white",
       main     = paste0(trait_name, ifelse(unit != "", paste0(" (", unit, ")"), "")),
       xlab     = trait_name,
       ylab     = "Count",
       xlim     = range(raw_vals, na.rm = TRUE),
       ylim     = c(0, max(hist(clean_vals, breaks=40, plot=FALSE)$counts) * 1.25),
       las      = 1,
       cex.main = 1.4,
       cex.lab  = 1.3,
       cex.axis = 1.3)

  if (any(outlier_mask, na.rm = TRUE))
    hist(raw_vals[outlier_mask],
         breaks = 40,
         col    = rgb(1, 0.3, 0.3, 0.6),
         border = NA,
         add    = TRUE)

  abline(v = trait_mean, col = "darkred", lwd = 2.5, lty = 2)

  legend("topright", bty = "n", cex = 1.3,
         text.col = c("black", "black", "red3", "darkred"),
         legend = c(
           paste0("N before: ", format(n_before, big.mark = ",")),
           paste0("N after:  ", format(n_after,  big.mark = ",")),
           paste0("Outliers removed: ", n_outliers),
           paste0("Mean: ", round(trait_mean, 2))
         ))
}

plot_int <- function(int_vals, trait_name) {
  n_final    <- sum(!is.na(int_vals))
  trait_mean <- mean(int_vals, na.rm = TRUE)
  trait_sd   <- sd(int_vals,   na.rm = TRUE)

  hist(int_vals,
       breaks   = 40,
       col      = "#2ecc71",
       border   = "white",
       main     = paste0(trait_name, " - after INT"),
       xlab     = "Inverse normal transformed value",
       ylab     = "Count",
       las      = 1,
       cex.main = 1.4,
       cex.lab  = 1.3,
       cex.axis = 1.3)

  abline(v = trait_mean, col = "darkgreen", lwd = 2.5, lty = 2)

  legend("topright", bty = "n", cex = 1.3,
         text.col = c("black", "darkgreen", "darkgreen"),
         legend = c(
           paste0("N: ",    format(n_final,    big.mark = ",")),
           paste0("Mean: ", round(trait_mean,  3)),
           paste0("SD: ",   round(trait_sd,    3))
         ))
}
```

### Raw distributions
```{r raw-distributions, fig.width=18, fig.height=12}
par(mfrow=c(2,3), mar=c(5,6.5,4,2), oma=c(1,1,2,1), mgp=c(4,1,0))
for (t in trait_info) plot_trait(t$raw, t$clean, t$name, t$unit)
mtext("Raw trait distributions (before cleaning)", outer=TRUE, cex=1.5, font=2)
```

### Post-INT distributions
```{r int-distributions, fig.width=18, fig.height=12}
par(mfrow=c(2,3), mar=c(5,6.5,4,2), oma=c(1,1,2,1), mgp=c(4,1,0))
for (t in trait_info) plot_int(t$int, t$name)
mtext("Trait distributions after covariate adjustment and INT", outer=TRUE, cex=1.5, font=2)
```

```{r save-dist-pdf}
pdf(file.path(output_dir, "phenotype_distributions.pdf"), width=18, height=12)

par(mfrow=c(2,3), mar=c(5,6.5,4,2), oma=c(1,1,2,1), mgp=c(4,1,0))
for (t in trait_info) plot_trait(t$raw, t$clean, t$name, t$unit)
mtext("Raw trait distributions (before cleaning)", outer=TRUE, cex=1.5, font=2)

par(mfrow=c(2,3), mar=c(5,6.5,4,2), oma=c(1,1,2,1), mgp=c(4,1,0))
for (t in trait_info) plot_int(t$int, t$name)
mtext("Trait distributions after covariate adjustment and INT", outer=TRUE, cex=1.5, font=2)

dev.off()
cat("Saved: phenotype_distributions.pdf\n")
```
---

## 12. SAIGE Output File

```{r saige-output}
saige_output <- phenotype_data %>%
  mutate(FID = genid, IID = genid) %>%
  select(FID, IID, bmi_INT, whr_bmi_adj_INT, waist_INT, hip_INT,
         fatmass_INT, fatperc_INT, ages, age_sq, sex_numeric, study)

cat("SAIGE output:", nrow(saige_output), "rows x", ncol(saige_output), "cols\n")

write.table(saige_output,
            file=file.path(output_dir, "believe_phenotypes_saige.txt"),
            sep="\t", row.names=FALSE, quote=FALSE, na="NA")
cat("Saved: believe_phenotypes_saige.txt\n")
```

---

## 13. Summary JSON for PowerPoint

```{r summary-json}

n_after_genid <- nrow(phenotype_data_raw %>% filter(!is.na(genid)))

data_quality_rows <- lapply(trait_info, function(t) {
  trait_key <- switch(t$name,
    "BMI"      = "bmi",
    "WHR"      = "whr",
    "Waist"    = "waist",
    "Hip"      = "hip",
    "Fat Mass" = "fatmass",
    "Fat %"    = "fatperc"
  )
  list(
    trait      = t$name,
    n_raw      = sum(!is.na(t$raw)),
    n_missing  = sum(is.na(t$raw)),
    n_negative = sum(t$raw < 0, na.rm = TRUE),
    n_outliers = outlier_summary$n_outliers[outlier_summary$trait == trait_key],
    n_final    = sum(!is.na(t$clean)),
    mean_final = round(mean(t$clean, na.rm = TRUE), 2),
    sd_final   = round(sd(t$clean,   na.rm = TRUE), 2)
  )
})

site_summary <- phenotype_data %>%
  group_by(study) %>%
  summarise(
    n            = n(),
    age_mean     = round(mean(ages,    na.rm=TRUE), 1),
    age_sd       = round(sd(ages,      na.rm=TRUE), 1),
    pct_female   = round(mean(sex=="Female", na.rm=TRUE)*100, 1),
    bmi_mean     = round(mean(bmi,     na.rm=TRUE), 2),
    bmi_sd       = round(sd(bmi,       na.rm=TRUE), 2),
    whr_mean     = round(mean(whr,     na.rm=TRUE), 3),
    whr_sd       = round(sd(whr,       na.rm=TRUE), 3),
    waist_mean   = round(mean(waist,   na.rm=TRUE), 1),
    waist_sd     = round(sd(waist,     na.rm=TRUE), 1),
    fatmass_mean = round(mean(fatmass, na.rm=TRUE), 1),
    fatmass_sd   = round(sd(fatmass,   na.rm=TRUE), 1),
    fatperc_mean = round(mean(fatperc, na.rm=TRUE), 1),
    fatperc_sd   = round(sd(fatperc,   na.rm=TRUE), 1)
  )

overall_summary <- phenotype_data %>%
  summarise(
    study        = "All sites",
    n            = n(),
    age_mean     = round(mean(ages,    na.rm=TRUE), 1),
    age_sd       = round(sd(ages,      na.rm=TRUE), 1),
    pct_female   = round(mean(sex=="Female", na.rm=TRUE)*100, 1),
    bmi_mean     = round(mean(bmi,     na.rm=TRUE), 2),
    bmi_sd       = round(sd(bmi,       na.rm=TRUE), 2),
    whr_mean     = round(mean(whr,     na.rm=TRUE), 3),
    whr_sd       = round(sd(whr,       na.rm=TRUE), 3),
    waist_mean   = round(mean(waist,   na.rm=TRUE), 1),
    waist_sd     = round(sd(waist,     na.rm=TRUE), 1),
    fatmass_mean = round(mean(fatmass, na.rm=TRUE), 1),
    fatmass_sd   = round(sd(fatmass,   na.rm=TRUE), 1),
    fatperc_mean = round(mean(fatperc, na.rm=TRUE), 1),
    fatperc_sd   = round(sd(fatperc,   na.rm=TRUE), 1)
  )

full_summary    <- bind_rows(site_summary, overall_summary)

summary_table_rows <- lapply(1:nrow(full_summary), function(i) {
  r <- full_summary[i,]
  list(
    study      = r$study,
    n          = r$n,
    age        = paste0(r$age_mean,     " (", r$age_sd,     ")"),
    pct_female = paste0(r$pct_female,   "%"),
    bmi        = paste0(r$bmi_mean,     " (", r$bmi_sd,     ")"),
    whr        = paste0(r$whr_mean,     " (", r$whr_sd,     ")"),
    waist      = paste0(r$waist_mean,   " (", r$waist_sd,   ")"),
    fatmass    = paste0(r$fatmass_mean, " (", r$fatmass_sd, ")"),
    fatperc    = paste0(r$fatperc_mean, " (", r$fatperc_sd, ")")
  )
})

n_male   <- sum(phenotype_data$sex == "Male",   na.rm = TRUE)
n_female <- sum(phenotype_data$sex == "Female", na.rm = TRUE)


summary_json <- list(

  total_participants = nrow(phenotype_data_raw),       
  after_genid_filter = n_after_genid,                  
  after_age_filter   = nrow(phenotype_data),           

  n_with_bmi         = sum(!is.na(phenotype_data$bmi)),
  n_with_fatmass     = sum(!is.na(phenotype_data$fatmass)),

  n_male             = n_male,
  n_female           = n_female,
  pct_male           = round(n_male   / nrow(phenotype_data) * 100, 1),
  pct_female         = round(n_female / nrow(phenotype_data) * 100, 1),

  summary_table      = summary_table_rows,
  data_quality       = data_quality_rows,

  sites = list(
    list(name="BELIEVE (Urban)",    n=sum(phenotype_data$study=="BELIEVE")),
    list(name="BELSLUM (Slum)",     n=sum(phenotype_data$study=="BELSLUM")),
    list(name="BELRURAL (Rural)",   n=sum(phenotype_data$study=="BELRURAL")),
    list(name="BELURBAN (Urban 2)", n=sum(phenotype_data$study=="BELURBAN"))
  ),


  traits = list(
    list(name="BMI",      unit="kg/m2",
         mean=round(mean(phenotype_data$bmi,     na.rm=TRUE), 2),
         sd  =round(sd(phenotype_data$bmi,       na.rm=TRUE), 2),
         n   =sum(!is.na(phenotype_data$bmi))),
    list(name="WHR",      unit="",
         mean=round(mean(phenotype_data$whr,     na.rm=TRUE), 3),
         sd  =round(sd(phenotype_data$whr,       na.rm=TRUE), 3),
         n   =sum(!is.na(phenotype_data$whr))),
    list(name="Waist",    unit="cm",
         mean=round(mean(phenotype_data$waist,   na.rm=TRUE), 1),
         sd  =round(sd(phenotype_data$waist,     na.rm=TRUE), 1),
         n   =sum(!is.na(phenotype_data$waist))),
    list(name="Hip",      unit="cm",
         mean=round(mean(phenotype_data$hip,     na.rm=TRUE), 1),
         sd  =round(sd(phenotype_data$hip,       na.rm=TRUE), 1),
         n   =sum(!is.na(phenotype_data$hip))),
    list(name="Fat Mass", unit="kg",
         mean=round(mean(phenotype_data$fatmass, na.rm=TRUE), 1),
         sd  =round(sd(phenotype_data$fatmass,   na.rm=TRUE), 1),
         n   =sum(!is.na(phenotype_data$fatmass))),
    list(name="Fat %",    unit="%",
         mean=round(mean(phenotype_data$fatperc, na.rm=TRUE), 1),
         sd  =round(sd(phenotype_data$fatperc,   na.rm=TRUE), 1),
         n   =sum(!is.na(phenotype_data$fatperc)))
  ),

  outliers_removed = list(
    list(trait="BMI",
         n=outlier_summary$n_outliers[outlier_summary$trait=="bmi"]),
    list(trait="WHR",
         n=outlier_summary$n_outliers[outlier_summary$trait=="whr"]),
    list(trait="Waist",
         n=outlier_summary$n_outliers[outlier_summary$trait=="waist"]),
    list(trait="Hip",
         n=outlier_summary$n_outliers[outlier_summary$trait=="hip"]),
    list(trait="Fat Mass",
         n=outlier_summary$n_outliers[outlier_summary$trait=="fatmass"]),
    list(trait="Fat %",
         n=outlier_summary$n_outliers[outlier_summary$trait=="fatperc"])
  )
)

write_json(summary_json, file.path(output_dir, "summary_stats.json"),
           pretty=TRUE, auto_unbox=TRUE)
cat("Saved: summary_stats.json\n")
```

---

## Summary

```{r final-summary}
cat("Output files saved to:", output_dir, "\n")
cat("  - believe_phenotypes_saige.txt\n")
cat("  - phenotype_distributions.pdf\n")
cat("  - corr_trait_to_trait.pdf\n")
cat("  - corr_cov_vs_raw.pdf\n")
cat("  - corr_cov_vs_int.pdf\n")
cat("  - boxplots_raw_by_site.pdf\n")
cat("  - boxplots_int_by_site.pdf\n")
cat("  - fatperc_by_sex.pdf\n")
cat("  - summary_stats.json\n")
```