---
title: "BELIEVE Phenotype Preparation"
author: "Joud Touqan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(jsonlite)
```

---

## 1. Read Data

```{r read-data}
phenotype_data_raw <- read.csv(
  "/home/jt962/rds/rds-post_qc_data-pNR2rM6BWWA/believe/phenotype/data_freeze_3/analysis_regeneron.csv",
  stringsAsFactors = FALSE)

cat("Number of participants:", nrow(phenotype_data_raw), "\n")
cat("Number of variables:",   ncol(phenotype_data_raw), "\n")
```

---

## 2. Select Columns

```{r select-columns}
phenotype_data <- phenotype_data_raw %>%
  select(genid, idno, study, ages, sex, ht, wt, bmi, waist, hip, whr, fatmass, fatperc)

cat("Columns selected:", nrow(phenotype_data), "rows x", ncol(phenotype_data), "columns\n")
```

---

## 3. Cleaning

```{r cleaning}
# a. Combine study + idno for a unique identifier across all sites
phenotype_data <- phenotype_data %>%
  mutate(unique_id = paste(study, idno, sep = "_"))

# b. Remove participants with missing geneid
n_without_genid <- sum(is.na(phenotype_data$genid))
phenotype_data <- phenotype_data %>%
  filter(!is.na(genid))
cat("Participants removed (Geneid = NA):", n_without_genid, "\n")
cat("Number of participants after filtering:", nrow(phenotype_data), "\n")

# c. Exclude children under 18
n_before_age <- nrow(phenotype_data)
phenotype_data <- phenotype_data %>%
  filter(ages >= 18)
cat("Participants removed (age < 18):", n_before_age - nrow(phenotype_data), "\n")
cat("Participants aged 18 and over (also full analysis cohort):", nrow(phenotype_data), "\n")

# d. Convert sex to numeric — SAIGE requires 0 = Male, 1 = Female
phenotype_data <- phenotype_data %>%
  mutate(sex_numeric = case_when(
    sex == "Male"   ~ 0,
    sex == "Female" ~ 1,
    TRUE            ~ NA_real_
  ))

cat("\nSex coding check:\n")
print(table(phenotype_data$sex, phenotype_data$sex_numeric, useNA = "always"))
```

---

## 4. Preliminary Exploration

### 4.1 Missing Data Summary

```{r missing-data}
missing_summary <- data.frame(
  trait     = c("ht", "wt", "bmi", "waist", "hip", "whr", "fatmass", "fatperc"),
  n_before  = sapply(c("ht", "wt", "bmi", "waist", "hip", "whr", "fatmass", "fatperc"),
                     function(v) sum(!is.na(phenotype_data[[v]]))),
  n_missing = sapply(c("ht", "wt", "bmi", "waist", "hip", "whr", "fatmass", "fatperc"),
                     function(v) sum(is.na(phenotype_data[[v]])))
)

missing_summary$n_after <- missing_summary$n_before - missing_summary$n_missing

missing_wide <- as.data.frame(t(missing_summary[, -1]))
colnames(missing_wide) <- missing_summary$trait
rownames(missing_wide) <- c("N with data", "N missing", "N after removing missing")
print(missing_wide)
```

### 4.2 Check for Impossible Values

```{r impossible-values}
cat("Number of impossible values:\n")
phenotype_data %>%
  summarise(
    bmi_neg     = sum(bmi     < 0, na.rm = TRUE),
    waist_neg   = sum(waist   < 0, na.rm = TRUE),
    hip_neg     = sum(hip     < 0, na.rm = TRUE),
    whr_neg     = sum(whr     < 0, na.rm = TRUE),
    fatmass_neg = sum(fatmass < 0, na.rm = TRUE),
    fatperc_neg = sum(fatperc < 0, na.rm = TRUE)
  ) %>% print()
```

---

## 5. Remove Outliers

Values > 5 SDs from the mean are set to NA. The participant is kept.

```{r outliers}
flag_outliers <- function(x) {
  mean_x  <- mean(x, na.rm = TRUE)
  sd_x    <- sd(x, na.rm = TRUE)
  outlier <- abs(x - mean_x) > 5 * sd_x
  outlier[is.na(x)] <- FALSE
  return(outlier)
}

traits <- c("bmi", "waist", "hip", "whr", "fatmass", "fatperc", "ht", "wt")

outlier_summary <- data.frame(
  trait      = traits,
  n_before   = NA_integer_,
  n_outliers = NA_integer_,
  n_after    = NA_integer_
)

for (i in seq_along(traits)) {
  trait    <- traits[i]
  n_before <- sum(!is.na(phenotype_data[[trait]]))
  outliers  <- flag_outliers(phenotype_data[[trait]])
  phenotype_data[[trait]][outliers] <- NA
  n_after  <- sum(!is.na(phenotype_data[[trait]]))

  outlier_summary$n_before[i]   <- n_before
  outlier_summary$n_outliers[i] <- n_before - n_after
  outlier_summary$n_after[i]    <- n_after
}

outlier_wide <- as.data.frame(t(outlier_summary[, -1]))
colnames(outlier_wide) <- outlier_summary$trait
rownames(outlier_wide) <- c("N before", "Outliers removed", "N after")
print(outlier_wide)
```

---

## 6. WHR Adjustment for BMI

WHR is regressed on BMI to capture fat distribution independent of overall body size. Residuals are used as adjusted WHR.

```{r whr-adjustment}
whr_model <- lm(whr ~ bmi, data = phenotype_data, na.action = na.exclude)
phenotype_data$whr_bmi_adj <- residuals(whr_model)
cat("WHR adjusted for BMI using linear regression.\n")
```

---

## 7. Covariate Adjustment

Age, age², sex, and study site are regressed out of each trait.

```{r covariate-adjustment}
phenotype_data <- phenotype_data %>%
  mutate(age_sq = ages^2)

adjust_trait <- function(trait_col, data) {
  formula <- as.formula(paste(trait_col, "~ ages + age_sq + sex_numeric + study"))
  model   <- lm(formula, data = data, na.action = na.exclude)
  resids  <- residuals(model)
  return(resids)
}

traits_to_adjust <- c("bmi", "waist", "hip", "whr_bmi_adj", "fatmass", "fatperc")

adjustment_summary <- data.frame(
  trait    = traits_to_adjust,
  n_before = NA_integer_,
  n_after  = NA_integer_
)

for (i in seq_along(traits_to_adjust)) {
  trait   <- traits_to_adjust[i]
  adj_col <- paste0(trait, "_adj")

  n_before <- sum(!is.na(phenotype_data[[trait]]))
  phenotype_data[[adj_col]] <- adjust_trait(trait, phenotype_data)
  n_after  <- sum(!is.na(phenotype_data[[adj_col]]))

  adjustment_summary$n_before[i] <- n_before
  adjustment_summary$n_after[i]  <- n_after
}

cat("Covariate adjustment complete. Summary:\n")
adj_wide <- as.data.frame(t(adjustment_summary[, -1]))
colnames(adj_wide) <- adjustment_summary$trait
rownames(adj_wide) <- c("N before adjustment", "N after adjustment")
print(adj_wide)
```

---

## 8. Inverse Normal Transformation (INT)

Adjusted traits are forced into a normal distribution using Blom's method.

```{r int-transformation}
int_transform <- function(x) {
  ranks       <- rank(x, na.last = "keep", ties.method = "average")
  n           <- sum(!is.na(x))
  transformed <- qnorm((ranks - 0.5) / n)
  return(transformed)
}

traits_adj <- paste0(traits_to_adjust, "_adj")

for (trait_adj in traits_adj) {
  int_col <- sub("_adj$", "_INT", trait_adj)
  phenotype_data[[int_col]] <- int_transform(phenotype_data[[trait_adj]])
}

cat("Inverse normal transformation applied to adjusted traits.\n")
```

---

## 9. Distribution Plots

```{r trait-info}
output_dir <- "/home/jt962/rds/hpc-work/believe_adiposity/01_phenotype_prep"

raw_18plus <- phenotype_data_raw %>% filter(ages >= 18)

trait_info <- list(
  list(raw = raw_18plus$bmi,     clean = phenotype_data$bmi,
       int = phenotype_data$bmi_INT,
       name = "BMI",      unit = "kg/m²", n_out = 30),
  list(raw = raw_18plus$whr,     clean = phenotype_data$whr,
       int = phenotype_data$whr_bmi_adj_INT,
       name = "WHR",      unit = "",      n_out = 20),
  list(raw = raw_18plus$waist,   clean = phenotype_data$waist,
       int = phenotype_data$waist_INT,
       name = "Waist",    unit = "cm",    n_out = 5),
  list(raw = raw_18plus$hip,     clean = phenotype_data$hip,
       int = phenotype_data$hip_INT,
       name = "Hip",      unit = "cm",    n_out = 29),
  list(raw = raw_18plus$fatmass, clean = phenotype_data$fatmass,
       int = phenotype_data$fatmass_INT,
       name = "Fat Mass", unit = "kg",    n_out = 19),
  list(raw = raw_18plus$fatperc, clean = phenotype_data$fatperc,
       int = phenotype_data$fatperc_INT,
       name = "Fat %",    unit = "%",     n_out = 0)
)
```

### 9.1 Plot Functions

```{r plot-functions}
plot_trait <- function(raw_vals, clean_vals, trait_name, unit, n_outliers) {
  outlier_mask <- flag_outliers(raw_vals)
  n_before     <- sum(!is.na(raw_vals))
  trait_mean   <- mean(clean_vals, na.rm = TRUE)

  hist(clean_vals,
       breaks   = 40,
       col      = "steelblue",
       border   = "white",
       main     = paste0(trait_name, ifelse(unit != "", paste0(" (", unit, ")"), "")),
       xlab     = trait_name,
       ylab     = "Count",
       xlim     = range(raw_vals, na.rm = TRUE),
       ylim     = c(0, max(hist(clean_vals, breaks = 40, plot = FALSE)$counts) * 1.25),
       las      = 1,
       cex.main = 1.4,
       cex.lab  = 1.3,
       cex.axis = 1.3)

  if (any(outlier_mask, na.rm = TRUE)) {
    hist(raw_vals[outlier_mask],
         breaks = 40,
         col    = rgb(1, 0.3, 0.3, 0.6),
         border = NA,
         add    = TRUE)
  }

  abline(v = trait_mean, col = "darkred", lwd = 2.5, lty = 2)

  legend("topright",
         legend = c(
           paste0("N before: ",        format(n_before, big.mark = ",")),
           paste0("Outliers removed: ", n_outliers),
           paste0("Mean: ",             round(trait_mean, 2))
         ),
         bty      = "n",
         cex      = 1.3,
         text.col = c("black", "black", "darkred"))
}

plot_int <- function(int_vals, trait_name, n_final) {
  trait_mean <- mean(int_vals, na.rm = TRUE)

  hist(int_vals,
       breaks   = 40,
       col      = "#2ecc71",
       border   = "white",
       main     = paste0(trait_name, " - after INT"),
       xlab     = "Inverse normal transformed value",
       ylab     = "Count",
       las      = 1,
       cex.main = 1.4,
       cex.lab  = 1.3,
       cex.axis = 1.3)

  abline(v = trait_mean, col = "darkgreen", lwd = 2.5, lty = 2)

  legend("topright",
         legend = c(
           paste0("N: ",    format(n_final, big.mark = ",")),
           paste0("Mean: ", round(trait_mean, 3)),
           paste0("SD: ",   round(sd(int_vals, na.rm = TRUE), 3))
         ),
         bty      = "n",
         cex      = 1.3,
         text.col = c("black", "darkgreen", "darkgreen"))
}
```

### 9.2 Raw Distributions

```{r raw-distributions, fig.width=18, fig.height=12}
par(mfrow = c(2, 3),
    mar   = c(5, 6.5, 4, 2),
    oma   = c(1, 1, 2, 1),
    mgp   = c(4, 1, 0))

for (t in trait_info) {
  plot_trait(t$raw, t$clean, t$name, t$unit, t$n_out)
}
mtext("Raw trait distributions (before cleaning)", outer = TRUE, cex = 1.5, font = 2)
```

### 9.3 Post-INT Distributions

```{r int-distributions, fig.width=18, fig.height=12}
par(mfrow = c(2, 3),
    mar   = c(5, 6.5, 4, 2),
    oma   = c(1, 1, 2, 1),
    mgp   = c(4, 1, 0))

for (t in trait_info) {
  plot_int(t$int, t$name, sum(!is.na(t$int)))
}
mtext("Trait distributions after covariate adjustment and INT", outer = TRUE, cex = 1.5, font = 2)
```

### 9.4 Save Distribution PDF

```{r save-pdf}
pdf(file.path(output_dir, "phenotype_distributions.pdf"), width = 18, height = 12)

par(mfrow = c(2, 3), mar = c(5, 6.5, 4, 2), oma = c(1, 1, 2, 1), mgp = c(4, 1, 0))
for (t in trait_info) { plot_trait(t$raw, t$clean, t$name, t$unit, t$n_out) }
mtext("Raw trait distributions (before cleaning)", outer = TRUE, cex = 1.5, font = 2)

par(mfrow = c(2, 3), mar = c(5, 6.5, 4, 2), oma = c(1, 1, 2, 1), mgp = c(4, 1, 0))
for (t in trait_info) { plot_int(t$int, t$name, sum(!is.na(t$int))) }
mtext("Trait distributions after covariate adjustment and INT", outer = TRUE, cex = 1.5, font = 2)

dev.off()
cat("Distribution plots saved to:", file.path(output_dir, "phenotype_distributions.pdf"), "\n")
```

---

## 10. SAIGE Output File

```{r saige-output}
saige_output <- phenotype_data %>%
  mutate(
    FID = genid,
    IID = genid
  ) %>%
  select(
    FID, IID,
    bmi_INT,
    whr_bmi_adj_INT,
    waist_INT,
    hip_INT,
    fatmass_INT,
    fatperc_INT,
    ages,
    age_sq,
    sex_numeric,
    study
  )

cat("SAIGE output dimensions:", nrow(saige_output), "rows x", ncol(saige_output), "cols\n")

out_saige <- file.path(output_dir, "believe_phenotypes_saige.txt")
write.table(
  saige_output,
  file      = out_saige,
  sep       = "\t",
  row.names = FALSE,
  quote     = FALSE,
  na        = "NA"
)
cat("SAIGE phenotype file saved to:", out_saige, "\n")
```

---

## 11. Summary JSON for PowerPoint

```{r summary-json}
# Step 1 — Data quality rows per trait
data_quality_rows <- lapply(trait_info, function(t) {
  list(
    trait      = t$name,
    n_raw      = sum(!is.na(t$raw)),
    n_missing  = sum(is.na(t$raw)),
    n_negative = sum(t$raw < 0, na.rm = TRUE),
    n_outliers = t$n_out,
    n_final    = sum(!is.na(t$clean)),
    mean_final = round(mean(t$clean, na.rm = TRUE), 2),
    sd_final   = round(sd(t$clean,   na.rm = TRUE), 2)
  )
})

# Step 2 — Site-level summary table
site_summary <- phenotype_data %>%
  group_by(study) %>%
  summarise(
    n            = n(),
    age_mean     = round(mean(ages,    na.rm = TRUE), 1),
    age_sd       = round(sd(ages,      na.rm = TRUE), 1),
    pct_female   = round(mean(sex == "Female", na.rm = TRUE) * 100, 1),
    bmi_mean     = round(mean(bmi,     na.rm = TRUE), 2),
    bmi_sd       = round(sd(bmi,       na.rm = TRUE), 2),
    whr_mean     = round(mean(whr,     na.rm = TRUE), 3),
    whr_sd       = round(sd(whr,       na.rm = TRUE), 3),
    waist_mean   = round(mean(waist,   na.rm = TRUE), 1),
    waist_sd     = round(sd(waist,     na.rm = TRUE), 1),
    fatmass_mean = round(mean(fatmass, na.rm = TRUE), 1),
    fatmass_sd   = round(sd(fatmass,   na.rm = TRUE), 1),
    fatperc_mean = round(mean(fatperc, na.rm = TRUE), 1),
    fatperc_sd   = round(sd(fatperc,   na.rm = TRUE), 1)
  )

# Step 3 — Overall summary row
overall_summary <- phenotype_data %>%
  summarise(
    study        = "All sites",
    n            = n(),
    age_mean     = round(mean(ages,    na.rm = TRUE), 1),
    age_sd       = round(sd(ages,      na.rm = TRUE), 1),
    pct_female   = round(mean(sex == "Female", na.rm = TRUE) * 100, 1),
    bmi_mean     = round(mean(bmi,     na.rm = TRUE), 2),
    bmi_sd       = round(sd(bmi,       na.rm = TRUE), 2),
    whr_mean     = round(mean(whr,     na.rm = TRUE), 3),
    whr_sd       = round(sd(whr,       na.rm = TRUE), 3),
    waist_mean   = round(mean(waist,   na.rm = TRUE), 1),
    waist_sd     = round(sd(waist,     na.rm = TRUE), 1),
    fatmass_mean = round(mean(fatmass, na.rm = TRUE), 1),
    fatmass_sd   = round(sd(fatmass,   na.rm = TRUE), 1),
    fatperc_mean = round(mean(fatperc, na.rm = TRUE), 1),
    fatperc_sd   = round(sd(fatperc,   na.rm = TRUE), 1)
  )

# Step 4 — Combine and convert to JSON-friendly list
full_summary <- bind_rows(site_summary, overall_summary)

summary_table_rows <- lapply(1:nrow(full_summary), function(i) {
  row <- full_summary[i, ]
  list(
    study      = row$study,
    n          = row$n,
    age        = paste0(row$age_mean,     " (", row$age_sd, ")"),
    pct_female = paste0(row$pct_female,   "%"),
    bmi        = paste0(row$bmi_mean,     " (", row$bmi_sd, ")"),
    whr        = paste0(row$whr_mean,     " (", row$whr_sd, ")"),
    waist      = paste0(row$waist_mean,   " (", row$waist_sd, ")"),
    fatmass    = paste0(row$fatmass_mean, " (", row$fatmass_sd, ")"),
    fatperc    = paste0(row$fatperc_mean, " (", row$fatperc_sd, ")")
  )
})

# Step 5 — Build full JSON
summary_json <- list(
  total_participants = nrow(phenotype_data_raw),
  after_age_filter   = nrow(phenotype_data),
  n_with_fatmass     = sum(!is.na(phenotype_data$fatmass)),
  n_with_bmi         = sum(!is.na(phenotype_data$bmi)),
  summary_table      = summary_table_rows,
  data_quality       = data_quality_rows,

  sites = list(
    list(name = "BELIEVE (Urban)",    n = sum(phenotype_data$study == "BELIEVE")),
    list(name = "BELSLUM (Slum)",     n = sum(phenotype_data$study == "BELSLUM")),
    list(name = "BELRURAL (Rural)",   n = sum(phenotype_data$study == "BELRURAL")),
    list(name = "BELURBAN (Urban 2)", n = sum(phenotype_data$study == "BELURBAN"))
  ),

  traits = list(
    list(name = "BMI",      unit = "kg/m\u00B2",
         mean = round(mean(phenotype_data$bmi,     na.rm = TRUE), 2),
         sd   = round(sd(phenotype_data$bmi,       na.rm = TRUE), 2),
         n    = sum(!is.na(phenotype_data$bmi))),
    list(name = "WHR",      unit = "",
         mean = round(mean(phenotype_data$whr,     na.rm = TRUE), 3),
         sd   = round(sd(phenotype_data$whr,       na.rm = TRUE), 3),
         n    = sum(!is.na(phenotype_data$whr))),
    list(name = "Waist",    unit = "cm",
         mean = round(mean(phenotype_data$waist,   na.rm = TRUE), 1),
         sd   = round(sd(phenotype_data$waist,     na.rm = TRUE), 1),
         n    = sum(!is.na(phenotype_data$waist))),
    list(name = "Hip",      unit = "cm",
         mean = round(mean(phenotype_data$hip,     na.rm = TRUE), 1),
         sd   = round(sd(phenotype_data$hip,       na.rm = TRUE), 1),
         n    = sum(!is.na(phenotype_data$hip))),
    list(name = "Fat Mass", unit = "kg",
         mean = round(mean(phenotype_data$fatmass, na.rm = TRUE), 1),
         sd   = round(sd(phenotype_data$fatmass,   na.rm = TRUE), 1),
         n    = sum(!is.na(phenotype_data$fatmass))),
    list(name = "Fat %",    unit = "%",
         mean = round(mean(phenotype_data$fatperc, na.rm = TRUE), 1),
         sd   = round(sd(phenotype_data$fatperc,   na.rm = TRUE), 1),
         n    = sum(!is.na(phenotype_data$fatperc)))
  ),

  outliers_removed = list(
    list(trait = "BMI",      n = sum(flag_outliers(phenotype_data_raw$bmi))),
    list(trait = "WHR",      n = sum(flag_outliers(phenotype_data_raw$whr))),
    list(trait = "Waist",    n = sum(flag_outliers(phenotype_data_raw$waist))),
    list(trait = "Hip",      n = sum(flag_outliers(phenotype_data_raw$hip))),
    list(trait = "Fat Mass", n = sum(flag_outliers(phenotype_data_raw$fatmass))),
    list(trait = "Fat %",    n = sum(flag_outliers(phenotype_data_raw$fatperc)))
  )
)

# Step 6 — Write JSON
json_path <- file.path(output_dir, "summary_stats.json")
write_json(summary_json, json_path, pretty = TRUE, auto_unbox = TRUE)
cat("JSON saved to:", json_path, "\n")
```

---

## Summary

```{r final-summary}
cat("Phenotype preparation done\n")
cat("Output files in:", output_dir, "\n")
cat("  - believe_phenotypes_saige.txt   (input for SAIGE GWAS)\n")
cat("  - phenotype_distributions.pdf    (distribution check plots)\n")
cat("  - summary_stats.json             (summary statistics for PowerPoint)\n")
```
